Topics in Algorithms

### 2018-01-10 project/paper.tex
lala

\section{Oracle with $O(\lg k)$ query time}
As aforementioned, binary search on the iteration sequence of the Thorup-Zwick oracle is performed to drive down the query time to $\lg k$. Starting with the iteration sequence $\mathcal{I}=1,\hdots,k-1$, the idea is to identify $r=O(\lg k)$ subsequences $(\mathcal{I}_1=\mathcal{I}) \supset \mathcal{I}_1 \supset \hdots \supset \mathcal{I}_r$, where for $2 \leq j \leq t$ we have $|\mathcal{I}_{j-1}| \geq 2|\mathcal{I}_j|$. We wish to maintain the invariant that $dist_k(u,v,i)$ applied to the first element of any subsequence $\mathcal{I}_j$ outputs the stretch $(2k-1)$ estimate in $|\mathcal{I}_j|$ time.
\\
\begin{defi}
For vertices $u,v \in V$ an index $j \in \mathcal{I}$ is $(u,v)$-terminal if
\begin{enumerate}
\item $j = k-1$ or
\item $j < k-1$ is even and $p_j(u) \in B_v$ or $p_{j+1}(v) \in B_u$ 
\end{enumerate}
We note that $j$ attains a $(u,v)$ terminal index, the algorithm terminates in the current or next iteration. \label{def:uvterm}
\end{defi}
The definition of $(u,v)$-terminality captures the essence of when $dist_k(u,v)$ terminates, i.e. when $p_j(u) \in B_v$. This happens either when 1) $j=k-1$, in which case, trivially, $p_{k-1}(u) \in B_v$ or 2) when $j < k-1$ is even and $p_j(u) \in B_v$ or $p_{j+1}(v) \in B_u$ (in this case $dist_k(u,v)$ terminates either in iteration $j$ or $j+1$), see Alg. \ref{alg:distk}.
\\
\begin{defi}
A subsequence $\mathcal{I'}=i_1, \hdots, i_2$ of $\mathcal{I}$ is $(u,v)$-feasible if
\begin{enumerate}
\item $i_1$ is even and
\item $d_{G}(p_{i_1}(u),u) \leq i_1 \cdot d_G(u,v)$ and
\item $i_2$ is $(u,v)$-terminal
\end{enumerate}
\label{def:uvfeas}
\end{defi}
Condition $1$ and $3$ of $(u,v)$-feasibility entails a symmetric view of $dist_k(u,v,i)$ in which $u$ and $v$ successively exhanges roles for even an uneven $j$. Condition $2$ ensures that $dist_k(u,v,i)$ returns a $(2k-1)$-approximation when run with the first  $(u,v)$-feasible subsequence, as seen in the following lemma:

\begin{lem}
Let $i_1, \hdots, i_2$ be a $(u,v)$-feasible subsequence. Then $dist_k(u,v,i_1)$ gives a $2k-1$ approximation of $d_G(u,v)$ in time $O(i_2-i_1)$.
\paragraph{Proof} By assumption $i_2$ is $(u,v)$-terminal. Therefore, $p_{i_2}(u) \in B_v$ or $p_{i_2 + 1}(v) \in B_u$. Thus $dist_k(u,v,i_1)$ terminates when $j=O(i_2)$ in time $O(i_2-i_1)$.

Furthermore $d_{G}(p_{i_1}(u),u) \leq i_1 \cdot d_G(u,v)$. When $p_j(u) = w \not\in B_v$
we have $d_G(v,p_{j+1}(v)) \leq d_G(v,p_{j}(u)) \leq d_G(v,u) + d_G(u,p_j(u))$ by Lemma \ref{lem:onedeltaperiteration} and the triangle inequality, as visualized in Fig. \ref{fig:j1}. Hence each iteration of $dist_k(u,v,i)$ increases $d_G(u,w)$ by at most $d_G(u,v)$. At termination we have
\begin{align*}
d_G(u,w) + d_G(w,v) &\leq 2d_G(u,w) + d_G(u,v) \\
&\leq 2(i_1 \cdot d_G(u,v) + (i_2 - i_1)d_G(u,v)) + d_G(u,v) \\
&\leq 2 \cdot i_2 \cdot d_G(u,v) + d_G(u,v) \\
&\leq (2 \cdot i_2 + 1)d_G(u,v) \\
&\leq (2 (k-1) + 1)d_G(u,v) \\
&\leq (2k-1) d_G(u,v)
\end{align*}
\end{lem}
% Inkluder illustration - forklarende tekst "By lemma 2.1 and the triangle inequality"

% Lemma 3.2 her - Hele sekvensen, svarende til at køre Thorup-Zwick i=1...k

% Lemma 3.3 her - "skip hver anden" - gør analysen symmetrisk - lav illustration (projicer punkter ned på reelle akse og indtegn afstande). Forklar bevis.

\begin{lem}
Let $i_1, \hdots, i_2$ be a $(u,v)$-feasible subsequence and let $i$ be even, $i_1+2 \leq i \leq i_2-2$. Let $j$ be an even index in subsequence $i_1, \hdots ,i-2$ that maximizes $\delta_j(u)$. If $p_j(u) \not\in B_v$ and $p_j+1(v) \not\in B_u$ then $i, \hdots , i_2$ is $(u,v)$-feasible. Otherwise, $i_1, \hdots , j$ is $(u,v)$-feasible.

\paragraph{Proof} If $p_j(u) \in B_v$ or $p_j+1(v) \in B_u$, then $j$ is $(u,v)$-feasible, and so is $i_1, \hdots, j$. Assume that $p_j(u) \not\in B_v$ and $p_j+1(v) \not\in B_u$. Then, by Lemma \ref{lem:onedeltaperiteration}, $d_G(v,p_{j+1}(v)) \leq d_G(v,p_{j}(u))$ and $d_G(u,p_{j+2}(u)) \leq d_G(u,p_{j+1(v)})$. Applying the triangle inequality twice, cfr. Fig. \ref{fig:twoiterations}, we get 
\begin{align*}
d_G(u,p_{j+2}(u) &\leq d_G(u,p_{j+1}(v)) &(\text{Lemma \ref{lem:onedeltaperiteration}}) \\
&\leq d_G(u,v) + d_G(v, p_{j+1}(v)) &(\text{Triangle Inequality})\\
&\leq d_G(u,v) + d_G(v, p_{j}(u)) &(\text{Lemma \ref{lem:onedeltaperiteration}})\\
&\leq 2d_G(u,v) + d_G(u, p_{j}(u))&(\text{Triangle Inequality})
\end{align*}
so $\delta_j(u) = d_G(u,p_{j+2}(u)) - d_G(u,p_j(u)) \leq 2d_G(u,v)$. Let $\mathcal{I}'$ be the set of even indices $i_1, i_1+2, i_1+4, \hdots, i-2$. Since $i_1, \hdots, i_2$ is $(u,v)$-feasible, $d_G(u,p_{i_1}(u)) \leq i_1 \cdot d_G(u,v)$. By the choice of $j$,
\begin{align*}
d_G(u,p_{i}(u)) &= d_G(u,p_{i_1}(u)) + \sum_{j' \in \mathcal{I'}}\delta_{j'}(u) \\
&\leq d_G(u,p_{i_1}(u)) + \left| \mathcal{I'} \right| \max_{j' \in \mathcal{I'}}\delta_{j'}(u) \\
&= d_G(u,p_{i_1}(u)) + \frac{i-i_1}{2} \delta_{j}(u) \\
&\leq i_1 \cdot d_G(u,v) + (i-i_1)d_G(u,v) \\
&= i \cdot d_G(u,v)
\end{align*}
Hence, since $i_1, \hdots, i_2$ is $(u,v)$-feasible, so is $i, \hdots, i_2$.  \label{lem:uvfeasibility}
\end{lem}

@@

\begin{figure}
\centering
    \includegraphics[width=1\textwidth]{include/tikz/distance_real_line.tikz}%
\caption{Visualization of the distance $d_G(u,p_j(u))$ and $\delta_j(u)$. The green lines correspond to the circles defined by $u$ and the radii $d_G(p_l(u))$ for $l \in \left\{ i_1, j, j+2, i-2, i, i_2 \right\}$.}
\end{figure}

\begin{figure}[H]
\begin{tikzpicture}[level distance=1.5cm,
  level 1/.style={sibling distance=3cm},
  level 2/.style={sibling distance=1.5cm}]
  \node[color=red] {$I'$}
    child[color=red] {node {$[0, ..., i]$}
    child[color=black,thick] {node[color=cyan] {$I^j$}}
      child[dashed] {node {$p(l_1)$}
      	child[solid, left=0.1cm, color=black] {node[color=cyan] {$l_1=[i_1, i_1+2]$}}
      	child[solid,color=black] {node[color=cyan] {$[i_1+2,i_1+4]$}}
      }
    }
    child[right=0.1cm, color=red] {node {$[i,...,k-1]$}
    child[color=black] {node[color=cyan] {$I^k$} 
    						child[dotted, color=black] {node {...}}
    						child[dotted, color=black,left=0.1cm] {node {...}}
    					}
      child[dashed] {{node {$p(l_2)$}
       	child[solid,color=black] {node[color=cyan] {$[i_2-4,i_2-2]$}}
      	child[right=1cm,solid,color=black] {node[color=cyan] {$l_2=[i_2-2, i_2]$}}
    }}};

@@

\begin{thm}
For an integer $k\geq 2$, a $(2k - 1)$-approximate distance oracle of G of size $O(kn^{1+1/k})$ and $O(\log k)$ query time can be constructed in $O\left( \min \left\lbrace kmn^{1+\frac{1}{k}}, \sqrt{k}m + kn^{1+c/k}\right\rbrace\right)$ time for some constant $c$.

\paragraph{Proof.} For variable $k\geq 3$, for each $u$ the bunch $B_u$ in computed in $O(kmn^{2})$ using \cite{Thorup2005}. The remaining preprocessing for $u$ is determining the necessary subsequences of $\mathcal{I}$ and the $j$ that maximized $\delta_j(u)$ for each subsequence.
The necessary subsequences are built by

@@

The recursion tree for $bdist_k$ for $u$ will have $\theta(k/\log k)$ children and $\theta(k/\log k)$ non-child nodes. Computing $j$ for each subsequence in $T$ will take $O(\log k)$ time per node, so total preprocessing time for $u$ is $O(k)$. 
\end{thm}

>>>>

\section{Oracle with $O(\lg k)$ query time}
As aforementioned, binary search on the iteration sequence of the Thorup-Zwick oracle is performed to drive down the query time to $\lg k$. Starting with the iteration sequence $\mathcal{I}=1,\hdots,k-1$, the idea is to identify $r=O(\lg k)$ subsequences $(\mathcal{I}_1=\mathcal{I}) \supset \mathcal{I}_1 \supset \hdots \supset \mathcal{I}_r$, where for $2 \leq j \leq t$ we have $|\mathcal{I}_{j-1}| \geq 2|\mathcal{I}_j|$. We wish to maintain the invariant that $dist_k(u,v,i)$ applied to the first element of any subsequence $\mathcal{I}_j$ outputs the stretch $(2k-1)$ estimate in $|\mathcal{I}_j|$ time.
\\
\begin{defi}
For vertices $u,v \in V$ an index $j \in \mathcal{I}$ is $(u,v)$-terminal if
\begin{enumerate}
\item $j = k-1$ or
\item $j < k-1$ is even and $p_j(u) \in B_v$ or $p_{j+1}(v) \in B_u$ 
\end{enumerate}
We note that $j$ attains a $(u,v)$ terminal index, the algorithm terminates in the current or next iteration. \label{def:uvterm}
\end{defi}
The definition of $(u,v)$-terminality captures the essence of when $dist_k(u,v)$ terminates, i.e. when $p_j(u) \in B_v$. This happens either when 1) $j=k-1$, in which case, trivially, $p_{k-1}(u) \in B_v$ or 2) when $j < k-1$ is even and $p_j(u) \in B_v$ or $p_{j+1}(v) \in B_u$ (in this case $dist_k(u,v)$ terminates either in iteration $j$ or $j+1$), see Alg. \ref{alg:distk}.
\\
\begin{defi}
A subsequence $\mathcal{I'}=i_1, \hdots, i_2$ of $\mathcal{I}$ is $(u,v)$-feasible if
\begin{enumerate}
\item $i_1$ is even and
\item $d_{G}(p_{i_1}(u),u) \leq i_1 \cdot d_G(u,v)$ and
\item $i_2$ is $(u,v)$-terminal
\end{enumerate}
\label{def:uvfeas}
\end{defi}
Condition $1$ and $3$ of $(u,v)$-feasibility employs a symmetric view of iterations of $dist_k(u,v,i)$ in which $u$ and $v$ successively exhanges roles for even an uneven $j$. Condition $2$ ensures that $dist_k(u,v,i)$ returns a $(2k-1)$-approximation when run with the first $(u,v)$-feasible subsequence, as seen in the following lemma:
\\
\begin{lem}
Let $i_1, \hdots, i_2$ be a $(u,v)$-feasible subsequence. Then $dist_k(u,v,i_1)$ gives a $2k-1$ approximation of $d_G(u,v)$ in time $O(i_2-i_1)$.
\paragraph{Proof} By assumption $i_2$ is $(u,v)$-terminal. Therefore, $p_{i_2}(u) \in B_v$ or $p_{i_2 + 1}(v) \in B_u$. Thus $dist_k(u,v,i_1)$ terminates when $j=O(i_2)$ in time $O(i_2-i_1)$.

Furthermore $d_{G}(p_{i_1}(u),u) \leq i_1 \cdot d_G(u,v)$. When $p_j(u) = w \not\in B_v$
we have $d_G(v,p_{j+1}(v)) \leq d_G(v,p_{j}(u)) \leq d_G(v,u) + d_G(u,p_j(u))$ by Lemma \ref{lem:onedeltaperiteration} and the triangle inequality. Hence each iteration of $dist_k(u,v,i)$ increases $d_G(u,w)$ by at most $d_G(u,v)$. At termination we have
\begin{align*}
d_G(u,w) + d_G(w,v) &\leq 2d_G(u,w) + d_G(u,v) \\
&\leq 2(i_1 \cdot d_G(u,v) + (i_2 - i_1)d_G(u,v)) + d_G(u,v) \\
&\leq 2 \cdot i_2 \cdot d_G(u,v) + d_G(u,v) \\
&\leq (2 \cdot i_2 + 1)d_G(u,v) \\
&\leq (2 (k-1) + 1)d_G(u,v) \\
&\leq (2k-1) d_G(u,v)
\end{align*}
\label{lem:uvfeasterminates}
\end{lem}

Furthermore the lemma shows that $dist_k(u,v,i)$ runs in time proportional to $i_2-i_1$. We also have the following ``base case'' for $\mathcal{I}$:
\\
\begin{lem}
$\mathcal{I}$ is $(u,v)$-feasible for all $u,v \in V$. \label{lem:basecase}
\end{lem}
The above lemma follows since $p_{k-1}(u) \in B_v$ for all $u \not= v \in V$. The following lemma gives us a way of splitting a $(u,v)$-feasible subsequence $I$ into two subsequences of size roughly $\frac{1}{2}|\mathcal{I}|$:
\\
\begin{lem}
Let $i_1, \hdots, i_2$ be a $(u,v)$-feasible subsequence and let $i$ be even, $i_1+2 \leq i \leq i_2-2$. Let $j$ be an even index in subsequence $i_1, \hdots ,i-2$ that maximizes $\delta_j(u)$. If $p_j(u) \not\in B_v$ and $p_j+1(v) \not\in B_u$ then $i, \hdots , i_2$ is $(u,v)$-feasible. Otherwise, $i_1, \hdots , j$ is $(u,v)$-feasible.

\paragraph{Proof} If $p_j(u) \in B_v$ or $p_j+1(v) \in B_u$, then $j$ is $(u,v)$-feasible, and so is $i_1, \hdots, j$. Assume that $p_j(u) \not\in B_v$ and $p_j+1(v) \not\in B_u$. Then, by Lemma \ref{lem:onedeltaperiteration}, $d_G(v,p_{j+1}(v)) \leq d_G(v,p_{j}(u))$ and $d_G(u,p_{j+2}(u)) \leq d_G(u,p_{j+1(v)})$. Applying the triangle inequality twice, cfr. Fig. \ref{fig:twoiterations}, we get 
\begin{align*}
d_G(u,p_{j+2}(u) &\leq d_G(u,p_{j+1}(v)) &(\text{Lemma \ref{lem:onedeltaperiteration}}) \\
&\leq d_G(u,v) + d_G(v, p_{j+1}(v)) &(\text{Triangle Inequality})\\
&\leq d_G(u,v) + d_G(v, p_{j}(u)) &(\text{Lemma \ref{lem:onedeltaperiteration}})\\
&\leq 2d_G(u,v) + d_G(u, p_{j}(u))&(\text{Triangle Inequality})
\end{align*}
so $\delta_j(u) = d_G(u,p_{j+2}(u)) - d_G(u,p_j(u)) \leq 2d_G(u,v)$. Let $\mathcal{I}'$ be the set of even indices $i_1, i_1+2, i_1+4, \hdots, i-2$. Since $i_1, \hdots, i_2$ is $(u,v)$-feasible, $d_G(u,p_{i_1}(u)) \leq i_1 \cdot d_G(u,v)$. By the choice of $j$,
\begin{align*}
d_G(u,p_{i}(u)) &= d_G(u,p_{i_1}(u)) + \sum_{j' \in \mathcal{I'}}\delta_{j'}(u) \\
&\leq d_G(u,p_{i_1}(u)) + \left| \mathcal{I'} \right| \max_{j' \in \mathcal{I'}}\delta_{j'}(u) \\
&= d_G(u,p_{i_1}(u)) + \frac{i-i_1}{2} \delta_{j}(u) \\
&\leq i_1 \cdot d_G(u,v) + (i-i_1)d_G(u,v) \\
&= i \cdot d_G(u,v)
\end{align*}
Hence, since $i_1, \hdots, i_2$ is $(u,v)$-feasible, so is $i, \hdots, i_2$.  \label{lem:uvfeasibility}
\end{lem}
Visualisations of the proof for the bound on $d_G(u,p_{j+2})$ across two iterations can be seen in Fig. \ref{fig:twoiterations}. A visualisation of the proof for the bound on $d_G(u,p_i(u))$ can be seen Fig. \ref{fig:realline}.

@@

\begin{figure}
\centering
    \includegraphics[width=1\textwidth]{include/tikz/distance_real_line.tikz}%
\caption{Visualization of the distance $d_G(u,p_j(u))$ and $\delta_j(u)$. The green lines correspond to the circles defined by $u$ and the radii $d_G(p_l(u))$ for $l \in \left\{ i_1, j, j+2, i-2, i, i_2 \right\}$. \label{fig:realline}}
\end{figure}

What lemmas \ref{lem:uvfeasterminates}, \ref{lem:basecase} and \ref{lem:uvfeasibility} entail is that, starting with the $(u,v)$-feasible sequence $\mathcal{I}=0,\hdots,k-1$, we may split it into two subsequences, one in which $(u,v)$-feasibility must be preserved. This splitting can be done recursively until a subsequence $\mathcal{I_r}$ for which $|\mathcal{I_r}|=O(\lg k)$ is determined. Applying $dist_k(u,v,i)$ to the first element of $\mathcal{I_r}$ produces a $(2k-1)$-approximation of the distance from $u$ to $v$ by $(u,v)$-feasibility in time proportional to $|\mathcal{I}_r|=O(\lg k)$.

\begin{figure}[h]
\begin{tikzpicture}[level distance=1.5cm,
  level 1/.style={sibling distance=3cm},
  level 2/.style={sibling distance=1.5cm}]
  \node[color=red] {$I'$}
    child[color=red] {node {$[0, ..., i]$}
    child[color=black,thick] {node[color=cyan] {$I^j$}}
      child[dashed] {node {$p(l_1)$}
      	child[solid, left=0.1cm, color=black] {node[color=cyan] {$l_1=[i_1, i_1+2]$}}
      	child[solid,color=black] {node[color=cyan] {$[i_1+2,i_1+4]$}}
      }
    }
    child[right=0.1cm, color=red] {node {$[i,...,k-1]$}
    child[color=black] {node[color=cyan] {$I^k$} 
    						child[dotted, color=black] {node {...}}
    						child[dotted, color=black,left=0.1cm] {node {...}}
    					}
      child[dashed] {{node {$p(l_2)$}
       	child[solid,color=black] {node[color=cyan] {$[i_2-4,i_2-2]$}}
      	child[right=1cm,solid,color=black] {node[color=cyan] {$l_2=[i_2-2, i_2]$}}
    }}};

@@

\begin{thm}
For an integer $k\geq 2$, a $(2k - 1)$-approximate distance oracle of G of size $O(kn^{1+1/k})$ and $O(\log k)$ query time can be constructed in $O\left( \min \left\lbrace kmn^{1+\frac{1}{k}}, \sqrt{k}m + kn^{1+c/k}\right\rbrace\right)$ time for some constant $c$.
\end{thm}
{\it Proof.} For variable $k\geq 3$, for each $u$ the bunch $B_u$ in computed in $O(kmn^{2})$ using \cite{Thorup2005}. The remaining preprocessing for $u$ is determining the necessary subsequences of $\mathcal{I}$ and the $j$ that maximized $\delta_j(u)$ for each subsequence.
The necessary subsequences are built

@@

The recursion tree for $bdist_k$ for $u$ will have $\theta(k/\log k)$ children and $\theta(k/\log k)$ non-child nodes. Computing $j$ for each subsequence in $T$ will take $O(\log k)$ time per node, so total preprocessing time for $u$ is $O(k)$.
